---
title: 'Model-based Inference in Ecology and Epidemiology'
date: ICTP Spring College 9--20 May 2016
output: html_document
bibliography: sbied.bib
csl: ecology.csl

---

```{r knitr-opts,include=FALSE,purl=FALSE,cache=FALSE}
library(knitr)
require(plyr)
require(reshape2)
require(magrittr)

prefix <- "syllabus"
opts_chunk$set(
  progress=TRUE,
  prompt=FALSE,tidy=FALSE,highlight=TRUE,
  strip.white=TRUE,
  warning=FALSE,
  message=FALSE,
  error=FALSE,
  echo=FALSE,
  cache=FALSE,
  cache.extra=rand_seed,
  results='markup',
  fig.show='asis',
  size='small',
  fig.lp="fig:",
  fig.path=paste0("figure/",prefix,"-"),
  cache.path=paste0("cache/",prefix,"-"),
  fig.pos="h!",
  fig.align='center',
  fig.height=4,fig.width=6.83,
  dpi=300,
  dev='png',
  dev.args=list(bg='transparent')
  )

options(
  keep.source=TRUE,
  stringsAsFactors=FALSE,
  encoding="UTF-8"
  )
```

### Instructor

Prof.&nbsp;Aaron A. King, Ph.D.  
Departments of Ecology & Evolutionary Biology and Mathematics  
University of Michigan  
Email: kingaa@umich.edu  

### Perspective

Ecological and epidemiological systems are particularly interesting from the physical point of view.
Their complexity and high-dimensionality makes it natural to approach them as stochastic, nonlinear dynamical systems and within this context, many questions of both intrinsic interest and practical concern can be formulated.
To answer these questions, it is necessary to rigorously confront hypothetical models with data.
In this regard, time-series data are of particular value inasmuch as they have the potential to express the characteristic signatures of causal mechanisms.
This course will take students into the heart of these issues via an introduction to ecological and epidemiological stochastic dynamical systems models using a series of examples with real data.
Students will learn how to formulate questions as models and answer the questions using state-of-the-art inference algorithms.

## Course objectives

1. to introduce partially observed Markov process (POMP) models as tools for scientific investigation
1. to give students the ability to formulate POMP models of their own
1. to teach efficient approaches for performing scientific inference using POMP models
1. to familiarize students with the **pomp** package
1. to give students opportunities to work with such inference methods
1. to provide documented examples for student re-use

### Prerequisites

- Familiarity with deterministic dynamics (discrete-time maps, ordinary differential equations) and probability.
- Some programming experience, in any language.
- Completion of the [**R** tutorial](http://kinglab.eeb.lsa.umich.edu/R_Tutorial) before the beginning of the course.
- A sense of humor.

### Format and expectations

The course will be taught using a mixture of lectures and computational laboratory exercises.
Students are expected to complete assigned readings before class meetings, keep up with assigned homework, participate fully in discussions, and work on course activities during class meetings.

### Course website

Materials and readings will be posted on the [course website](http://kingaa.github.io/short-course).

### Computing in **R**

We will make extensive use of the open-source **R** statistical computing environment and the **pomp** package for inference based on partially-observed Markov process models.

Students with laptops should install **R** and **RStudio** on their computers before the first day of the course.
[Instructions for doing so can be found here](./prep/preparation.html).

The course does not assume familiarity with **R**, but students should work through the [**R** tutorial](http://kinglab.eeb.lsa.umich.edu/R_Tutorial) before the course commences.
In particular, students should work through the the exercises in the tutorial.

### Readings

The following papers serve as background for some of the central issues:

- S. N. Wood (2010) Statistical inference for noisy nonlinear ecological dynamic systems. *Nature*, **466**:&nbsp;1102&ndash;1104. [DOI:&nbsp;10.1038/nature09319](http://dx.doi.org/10.1038/nature09319).
- A. A. King, E. L. Ionides, M. Pascual, and M. J. Bouma (2008) Inapparent infections and cholera dynamics. *Nature*, **454**:&nbsp;877&ndash;880. [DOI:&nbsp;10.1038/nature07084](http://dx.doi.org/10.1038/nature07084)
- S. Shrestha, A. A. King, and P. Rohani (2011) Statistical Inference for Multi-Pathogen Systems. *PLoS&nbsp;Comput.&nbsp;Biol.*, **7**:&nbsp;e1002135. [DOI:&nbsp;10.1371/journal.pcbi.1002135](http://dx.doi.org/10.1371/journal.pcbi.1002135)

A good reference for modeling in infectious disease epidemiology is:

- Keeling, M. & Rohani, P. *Modeling infectious diseases in humans and animals*. Princeton University Press, 2008.

The **pomp** package is described and illustrated in

- A. A. King, D. Nguyen, and E. L. Ionides (2016) Statistical Inference for Partially Observed Markov Processes via the R Package pomp. *J.&nbsp;Stat.&nbsp;Soft.*, **69**:&nbsp;1&ndash;43. [DOI:&nbsp;10.18637/jss.v069.i12](http://dx.doi.org/10.18637/jss.v069.i12)

### Tentative schedule

```{r sched1,echo=FALSE}
begin.date <- as.Date("2016-05-06")
data.frame(date=begin.date+cumsum(c(rep(c(3,1,1,1,1),times=2)))) -> dates

read.csv(colClasses=c(Date="Date"),comment.char="#",
         text='
Date,Topic
2016-05-09,Ecological and epidemiological dynamics; introduction to **R**
2016-05-10,Partially-observed Markov processes; introduction to **pomp**
2016-05-11,Deterministic dynamics; trajectory matching
2016-05-12,Stochastic simulation
2016-05-13,Likelihood for POMP models
2016-05-16,Iterated filtering
2016-05-17,Case study: influenza in a closed population
2016-05-18,Diagnostics
2016-05-19,Case study: measles and extrademographic stochasticity
2016-05-20,Case study: pertussis and natural immune boosting
') %>%
  mutate(Topic=mapvalues(as.character(Topic),NA,""),
         Date=format(Date,"%b %e")) %>% 
  kable()
```

----------------------

## Schedule of Topics

1. [Ecological and epidemiological dynamics; introduction to **R**]&nbsp;&nbsp;&nbsp;[(**R** script)](http://raw.githubusercontent.com/kingaa/sbied/gh-pages/intro/intro.R)
1. Deterministic dynamics; trajectory matching
1. [Stochastic simulation of dynamic models.](./stochsim/stochsim.html)&nbsp;&nbsp;&nbsp;[(**R** script)](http://raw.githubusercontent.com/kingaa/sbied/gh-pages/stochsim/stochsim.R)
1. [Likelihood for POMP models: theory and practice](./pfilter/pfilter.html)&nbsp;&nbsp;&nbsp;[(**R** script)](http://raw.githubusercontent.com/kingaa/sbied/gh-pages/pfilter/pfilter.R)
1. [Iterated filtering: theory and practice](./mif/mif.html)&nbsp;&nbsp;&nbsp;[(**R** script)](http://raw.githubusercontent.com/kingaa/sbied/gh-pages/mif/mif.R)
1. Case study: influenza in a closed population
1. [Case study: the persistence of polio.](./polio/polio.html)&nbsp;&nbsp;&nbsp;[(**R** script)](http://raw.githubusercontent.com/kingaa/sbied/gh-pages/polio/polio.R)
1. [Case study: measles and extra-demographic stochasticity.](./measles/measles.html)&nbsp;&nbsp;&nbsp;[(**R** script)](http://raw.githubusercontent.com/kingaa/sbied/gh-pages/measles/measles.R)
1. [Case study: Ebola and forecasting](./ebola/ebola.html)&nbsp;&nbsp;&nbsp;[(**R** script)](http://raw.githubusercontent.com/kingaa/sbied/gh-pages/ebola/ebola.R)

----------------------

[**pomp** homepage](http://kingaa.github.io/pomp)  

[Source code for these notes](http://github.com/kingaa/short-course/)  

----------------------
